---
title: "Important functions for data wrangling"
author: "Template: Ian Hussey; Content: André Minder"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: hide
    highlight: haddock
    theme: flatly
    toc: yes
    toc_float: yes
---

```{r, include=FALSE}

# set knit options
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE)

# disable scientific notation
options(scipen = 999) 

```

# Instructions

Most data wrangling tasks can be accomplished with a relatively small number of functions. I've listed the most important ones here. 

Use this file to keep notes about what a given function does in your own words, the situations where you would use it, and working examples. You can make use of built-in datasets to do this (e.g., `mtcars`) or load external data sets (although its easier to break the examples if you modify the data etc.).

Learning how to wrangle data efficiently involves a combination of knowing how to break the problem down into smaller components; knowing which functions are available to accomplish each component operation and how to use them; how to search documentation to learn or refresh your knowledge of how a function works; and how to discover new functions that accomplish new component operations. 

Remember that you can look up the help documentation for any function by typing a question mark followed by its name in the console, e.g., `?dplyr::mutate`. The help documentation provides details of a function's parameters and defaults, its outputs, and examples of its use. Note that when you can also open the help documentation for an entire package by typing the package name, e.g., `?dplyr`. This can be very useful to discover what other functions that package has: scroll down to the bottom of any help page and click the "Index" link to see all help pages for that package.

I have prepended each function below with package it comes from so that you know. For example, `summarize` is listed as `dplyr::summarize`. Usually you don't have to do this when using a function, although you can use this to resolve a common bug known as name conflicts (see [this blog post](https://www.r-bloggers.com/2010/08/namespaces-and-name-conflicts/) for discussion). 

# Resources (for this session and others)

- You can find cheatsheets for the dplyr, tidyr, and RMarkdown in the /resources/cheatsheets folder.
- The Open Source textbook R for Data Science (aka, [Wickham's R4DS](https://r4ds.hadley.nz/)) is invaluable. Hadley Wickham is the main developer of the "tidyverse" set of packages, including dplyr, tidyr, ggplot2, stringr, lubridate, and others. See its [section on data transformation](https://r4ds.hadley.nz/data-transform). 
  - The entire second edition of the book is available at [https://r4ds.hadley.nz/](https://r4ds.hadley.nz/).
  - The first edition is also available. It does some things better in my opinion, e.g., it has a better explanation of the pipe (`%>%` or `|>`). See [https://r4ds.had.co.nz/pipes.html](https://r4ds.had.co.nz/pipes.html). 
  - The first edition also talks about RMarkdown, whereas the second edition has moved to a different technology called Quarto (which we won't cover, although they're similar). See [https://r4ds.had.co.nz/r-markdown.html](https://r4ds.had.co.nz/r-markdown.html).
- For those of you who prefer to learn in an interactive environment, I now suggest this web app over Swirl as it is more user-friendly: [https://allisonhorst.shinyapps.io/dplyr-learnr/#section-welcome](https://allisonhorst.shinyapps.io/dplyr-learnr/#section-welcome).
- For those of you who prefer some video content - although seeing other people code can never replace practicing coding yourself! - I can also recommend De Bruine et al.'s Open Source textbook and videos [Data Skills for Reproducible Research](https://psyteachr.github.io/reprores-v3/). E.g., see their page with links to videos for [dplyr](https://psyteachr.github.io/reprores-v3/dplyr.html) and [tidyr](https://psyteachr.github.io/reprores-v3/tidyr.html). 

# Dependencies

The packages these functions come from, which must be loaded to use them.

```{r}

library(readr)
library(dplyr)
library(tidyr)
library(knitr)
library(kableExtra)
library(janitor)

```

# Basics

## read_csv 

vs. `read.csv()`

*Always, always, always* use relative paths rather than absolute paths.

- Absolute path (bad): "~/Ian/Desktop/my_project/data/data_raw.csv"
- Relative path (good): "../data/data_raw.csv"

When using relative paths: "../" means up one directory level, whereas "directory_name/" goes down one directory level.

Relative paths will work on other machines, where absolute paths will break. Note that relative paths only work in .Rmd files and not .R files. Even without RMarkdown's other benefits, this makes them worth using.

Useful arguments:

- 'skip' can be used to skip the first n rows when reading in a file

```{r}

# load the data (with a relative path), skipping the first 2 rows
# data <- readr::read_csv("../../data/AMP study/raw/data_demographics_raw_messy.csv", skip = 2)

```

## dir.create

The function 'dir.create' creates a new folder/directory at a specified location. This is useful if we need to establish a folder structure for someone else on a different computer. We even can produce whole folder structure from scratch. This ultimately contributes to the codes and projects reproducibility. 

```{r}

dir.create("results") # creates a folder called 'results' 

```

## colnames & dput

The function 'colnames' is an easy way to get the names/headers of columns from a df. 

```{r}

column_headers <- colnames(mtcars) # saves column headers in a new variable

# put(colnames(mtcars))

```

# Wrangling

## the pipe: `%>%` or `|>`

The pipe can be used for channeling the output of a function into another function. This supports code readability (because it is easily comprehensible how functions are nested) and helps avoiding errors.

There is an old version of the pipe ( %>% ) and a new one ( |> ). The old pipe was part of the magrittr/dplyr package, while the new pipe is part of base R. This makes it possible to use the pipe without loading any packages. Note that the new version of the pipe is not compatible with all functions.

```{r}

mtcars_new_v1 <- colnames(rename(slice(mtcars, 2:n()), ghp = hp)) # this version of code is hard to read

mtcars_new_v2 <- mtcars |>      # this version of code is much easier to read (and to write!). Note that both  
  slice(2:n()) |>               # version arrive at the same result.
  rename(ghp = hp) |> 
  colnames()

```

## round_half_up

The function 'round_half_up' from the 'janitor' rounds numbers in a more intuitive way than the base R function 'round'. The function 'round' performs the so-called bankers rounding, which rounds up or down not only depending on the numbers after the decimal point, but also depending on the number before the decimal point. In contrast, 'round_half_up' rounds numbers 'normally' (up if => .5, down if < .5).

```{r}

vector <- c(0.5, 1.5, 2.5, 3.5, 4.5)

# Bankers rounding in action
round(vector)

# More intuitive rounding
round_half_up(vector)

```

## clean_names

The function 'clean_names' from the 'janitor' package cleans up unhandy variable names (e.g. variable names with special characters) and renames them appropriately. This is helpful for large data sets with many variables, because other functions like 'rename' ask you to specify each target variable and its new name.

```{r}


mtcars_bad_headers <- mtcars |>     # This code generates a data set with inconvenient headers. Note that
  rename(`! mpg` = mpg,             # if we would rename the headers back to their original name, we would 
         `.cyl` = cyl,              # need quite some lines of code
         `Gear` = gear)

mtcars_good_headers <- janitor::clean_names(mtcars_bad_headers) # Column headers are fixed with one line of code!

```

## filter

The function 'filter' from the 'dplyr' package selects rows according to some specified condition (e.g. if the rows have a value higher than x in a variable, or a certain character value). Conditions can be constructed with logical operators (e.g. <, <=, !=). Also, several conditions can be linked within one filter function (e.g. &, |)

Useful helper functions:

- '%in%' helps selecting a vector of values from a column (e.g. several categories)

```{r}

# only select rows of the iris species 'setosa'
iris_setosa <- iris |>                              
  filter(Species == "setosa")

# only select rows with flowers that have a sepal widsmallerthan 2.5 and belong to the species 'setosa' (linked conditions) 
iris_setosa_small_sw <- iris |>                     
  filter(Sepal.Width < 2.5 & Species == "setosa")   

# Select the species 'setosa' and 'versicolor' from the data set
iris_setosa_versicolor <- iris |> 
  filter(Species %in% c("setosa", "versicolor"))    

```

## slice

The function 'slice' is retaining or dropping rows from a df. It might be used just after reading in files if there are junk header rows.

```{r}

mtcars_new <- mtcars |>     # Select the data from row 2 to the last row
  slice(2:n())

```

## select

The function 'select' from the 'dplyr' package makes it possible to select columns from a df. Note that several selection features can be combined with the symbols & and |. There are also several helper functions that help select the needed columns.

Useful helper functions:

- 'everything()' selects all columns - this might be helpful when columns need to be reordered (see below)
- 'starts_with()' selects the columns starting with certain characters (not case-sensitive)
- 'ends_with()' selects the columns ending with certain characters
- 'contains()' selects the columns that contain certain characters

```{r}

# Select the first 5 columns but also omit the column for the variable 'cyl'
mtcars_small <- mtcars |>   
  select(1:5 & !cyl)

# Only select the columns that start with 'petal' and the species column
iris_petal <- iris |> 
  select(starts_with("petal"), Species)   

 # only select the columns that end with 'width' and the species column
iris_width <- iris |> 
  select(ends_with("width"), Species)    

# Rearrange the iris data set by first selecting the column 'Species' and then all the other columns
iris_rearranged <- iris |>          
  select(Species, everything())

```

## rename

The function 'rename' from the 'dplyr' package can be used to rename column headers. Note that using the function in combination with larger data sets might be an arduous process, since every target variable needs to be listed separately. It is better to use the function 'clean_names' in this case.

```{r}

# Rename the headers of the iris data set to facilitate subsequent coding and avoid time-consuming mistakes (e.g. getting rid of capital letters)
iris_renamed <- iris |>                   
  rename(sepal_length = Sepal.Length,     
         sepal_width = Sepal.Width,
         petal_length = Petal.Length,
         petal_width = Petal.Width,
         species = Species)

```

## mutate

The function 'mutate' from the 'dplyr' package makes it possible to create new variables within a df. This is helpful if we want to combine variables (e.g. to calculate a difference score) or if variables are needed in a different unit.

Useful arguments:

- '.after' inserts the manipulated columns after a specified column (and not at the right side of the df)
- '.before' inserts the manipulated columns before a specified column (and not at the right side of the df)
- '.unused' drops the columns used for creating the new columns

```{r}

# create a new variable 'kg' (kilogram) in the df based on the variable 'wt' (weight) which contains values with the unit lbs
mtcars_kg <- mtcars |>        
  mutate(kg = wt * 453.592)

# Create two new variables in the iris data set that describe the ratio between petal/sepal width/length.
# Note the argument .after at the end, which inserts the columns after the column 'Petal.Width'.
iris_sepal_petal_ratio <- iris |> 
  mutate(sepal_petal_ratio_width = Sepal.Width / Petal.Width,            
         sepal_petal_ratio_length = Sepal.Length / Petal.Length,         
         .after = Petal.Width)

```

## case_when

The function 'case_when' from the 'dplyr' can be used to modify data according to several specified conditions. This is of great use, if we need to homogenize inputted data (e.g. gender) so we can summarize it more easily.
```{r}

# Making the iris data set a bit messy with case_when
iris_messy <- iris |> 
  mutate(Species = case_when(Species == "setosa" ~ toupper("setosa"),
                             Species == "versicolor" ~ "VerSicolor",
                             Species == "virginica" ~ Species))

# Unmessy it with case_when
iris_cleaned <- iris_messy |> 
  mutate(Species = case_when(Species == "virginica" ~ Species,
                             Species == "SETOSA" ~ tolower("SETOSA"),
                             Species == "VerSicolor" ~ "versicolor"))
  

```

## summarize

The function 'summarize' from the 'dpylr' package can be used to aggregate values and calculate summarize statistics.

Useful arguments:

- '.by' can be used to get the summary statistics according to some groups or categories (similar to function 'group_by')
```{r}

# Calculate the median, mean, sd, max and min of Sepal.Length
iris_stats <- iris |> 
  summarise(
    n = n(),
    iris_median = quantile(Sepal.Length, .5, na.rm = TRUE),
    iris_mean = mean(Sepal.Length, na.rm = TRUE),
    iris_sd = sd(Sepal.Length, na.rm = TRUE),
    iris_max = max(Sepal.Length, na.rm = TRUE),
    iris_min = max(Sepal.Length, na.rm = TRUE),
    .by = "Species"
  )


```

## count

The function 'count' from the 'dplyr' package counts how many times a category (e.g. iris setosa) appears within a variable (e.g. species). This helps to quickly summarize a data set and is very useful if we need to check if a variable has messy/faulty inputs (e.g. 'woman' instead of 'female' in a gender variable). Conceptually, 'count' does the same as grouping a data set according to a variable and then using summarize to get the n of the respective category. 

```{r}

# First variant using group_by and summarize to counts how many times 
# each species appears in the variable 'species'
iris_groups_v1 <- iris |> 
  group_by(Species) |> 
  summarize(n = n())

#  Simpler variant with count
iris_groups_v2 <- iris |>  
  count(Species)

```

## distinct

The function 'distinct' from the 'dplyr' package excludes rows that are exact duplicates of each other. It thus helps finding and getting rid of duplicates. In general, it is important to check the number of certain important values (e.g. participants) before and after the wrangling, so one can make sure that nothing relevant was lost during the process.

```{r}

# duplicating the whole data set
iris_duplicates <- iris |> 
  bind_rows(iris)


# Getting rid of the duplicates (and one additional row that had the same values as another row in the original data set -> Pay attention when using this function!)
iris_cleaned <- iris_duplicates |>   
  distinct()

```

## group_by

The function 'group_by' from the 'dplyr' package groups a data set according to a specified variable. This is helpful if we want to summarize or visualize data in relation to some groups or categories. We can also use the argument '.by' within the 'summarize' function to achieve the same.

```{r}

# Getting the n of each iris species
iris_species_v1 <- iris |> 
  group_by(Species) |> 
  summarize(n = n())

# The same as above, but with the argument .by
iris_species_v2 <- iris |> 
  summarize(n = n(),
            .by = Species)

```

## rowwise

'Rowwise' is a function from the 'dpylr' package. 'Rowwise' makes it possible to use operations (like taking the mean) across several rows.

```{r}

iris_row_means <- iris |> 
  select(Sepal.Length, Petal.Length) |> 
  rowwise() |> 
  mutate(length_row_means = mean(c(Sepal.Length, Petal.Length)))
  

```

## lead

The function 'lead' from the 'dplyr' package introduces NA values at the end of a vector or column of a df.

Useful arguments:

- n: Number of NA values to be introduced
```{r, results='hide'}

# Introducing two NA value at the end of the vector. Importantly, the first two values disappear!
lead(1:10, n = 2) 

# Introducing NA values to the column 'species' in the iris data set
iris_NAs <- iris |>  
  mutate(Species = lag(Species, n = 3)) |> 
  arrange(Species)

```

## lag


The function 'lag' from the 'dplyr' package introduces NA values at the start of a vector or column of a df.

Useful arguments:

- n: Number of NA values to be introduced
```{r, results='hide'}

# Introducing four NA values at the beginning of the vector
lag(1:10, n = 4)

```

## joins

The join functions from the 'dplyr' package are useful for merging data sets. They have a clear advantage over functions like 'cbind', because they combine dfs in a reproducible and easy way, making sure information is related properly across variables. There are several different join functions:

- full_join: Saves all information of both data sets in one data set
- left_join: Keeps all the data from the left df, while not keeping the unmatched data of the right data frame
- right_join: Keeps all the data from the right df, while not keeping the unmatched data of the left data frame
- inner_join: Keeps only the matched data between both dfs (i.e. the intersect data)
- semi_join: Keeps all the data from the left df, while excluding all the data from the right df plus the intersect data of the two dfs (this might be handy if we join data with an exclusion df)

Useful arguments:

- 'by' specifies which variables/keys are matched across dfs. 'By' = x is a short form of 'join_by(x == x), meaning that we assume that the key variable is equal in both dfs.

```{r}

# iris data set with different values
iris_new <- iris |> 
  mutate(Sepal.Length_t2 = Sepal.Length * 0.5,
         Sepal.Width_t2 = Sepal.Width * 0.4) |> 
  slice(1:100)
  
# Save the complete information of both dfs in one df
all_data <- full_join(iris, iris_new)

# Keep all the data from the iris data set and only keep the matched data from the iris_new data set
left_data <- left_join(iris, iris_new, by = c("Species", "Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width"))

# Keep all the data from the iris_new data set and only keep the matched data from the iris data set
right_data <- right_join(iris, iris_new, by = c("Species", "Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width"))

# Only keep the matched data of both dfs (intersect)
inner_data <- inner_join(iris, iris_new, by = c("Species", "Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width"))

# Only keep the data of the left df, excluding all data from the right df and the intersection of the left and right df
semi_data <- semi_join(iris, iris_new, by = c("Species", "Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width"))

```

## pivots

The pivot functions from the 'tidyr' package help changing the dimension or format of a data set. 'Pivot_longer' can transform a data set into a longer data set (meaning a data set with less columns and more rows), while 'pivot_wider' makes a data set wider (meaning a data set with more columns and less rows). This is of help if we need to have the data set in a specific format (e.g. for visualization with ggplot).

```{r}

iris_longer <- iris |> 
  pivot_longer(cols = !Species,
               names_to = c("flower_part", "dimension"),
               names_sep = "\\.",
               values_to = "values"
               )

iris_wider <- iris_longer |> 
  pivot_wider(
    names_from = c("flower_part", "dimension"),
    values_from = "values"
    )

```

## drop_na

The function 'drop_na' ('tidyr' package) drops rows which contain NA values.

```{r}

# Generate a data set with NA values
iris_na <- iris |> 
  mutate(Sepal.Length = replace(Sepal.Length, Sepal.Length == 5, NA))

# Drop rows with NA values
iris_noNa <- iris_na |> 
  drop_na()

```

## separate

```{r}

# tidyr::separate()

```

## fill

```{r}

# tidyr::fill()

```

## lapply

'lapply' is a base R function that applies a function to a set of variables. This is very handy if we need to know the types of all variables in our data sets quickly.
```{r}

lapply(iris, class)

```

# Printing tables

The functions 'kable' (package 'knitr') and 'kable_classic' (package 'kableExtra') can be used to generate and print nice-looking tables.

```{r}

mtcars |> # example data
  kable() |> # print a nicer looking table
  kable_classic(full_width = FALSE) # print nicer again

```

# Other packages 

Other packages you may need for wrangling which aren't covered here:

- library(forcats). Everything to do with factors and factor levels. Useful for plotting and establishing reference levels for statistical tests.
- library(stringr). Everything to do with strings, searching for strings, modifying strings, parsing them, etc.
- library(lubridate). Everything to do with dates, parsing dates, converting date formats, etc. 


# Helpful GitHub basics

## gitignore
Gitingore is a feature of GitHub. It "hides" selected files in a sense that they are not visible in the commits or pushed to the repository. This allows us to work locally on a data set, while preventing sensible information from spreading online. It makes sense to filter out sensible information before the first commit (and ignoring the data set with the sensible information), because the commits are also visible online. Also, there is the danger of abandoning made changes and reverting back to a version, where the data set still contains the sensible information.

# Session info

```{r}

sessionInfo()

```



