---
title: "Plot assignment"
author: "Template: Ian Hussey; content: [Student name]"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: hide
    highlight: haddock
    theme: flatly
    toc: yes
    toc_float: yes
---

```{r, include=FALSE}

knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE)

```

# Dependencies

```{r}

library(tidyverse)
library(faux)
library(janitor)

```

# Instructions

The goal of this exercise is to produce both good and bad examples of plots *and to be explicit about what makes them good or bad*. 

The data can be real or fabricated. The plots can convey realistic or silly messages - but they must convey them clearly. You can get real data from many sources. A good one is the open science framekwork (osf.io). You can also simply make up data, either by hand or using simulation functions (e.g., `rnorm`) or the many simulation packages. If simulating data, ensure that you (a) use set.seed() to make it reproudbile, and (b) save a copy of the final data that you're using to disk and then reload it in the script so that the plot is reproducible.

The goal is neither to reproduce the types of plots that you most often see in journals (bar plots, etc.) nor to use a very fancy looking plot just for the sake of it - if it is very unlikely that you'd ever be able to use the plot to convey a research finding in an article, blog, poster, tweet, infographic or otherwise, it's probably not a good choice either.

The plots must be reproducible. Provide the data as well as the code to reporduce them. The plots should also be saved as individual images (pdf or png).

Under each plot, here in the RMarkdown, briefly explain what types of analysis the plot is suitable for, what elements of the results the plot attempts to clearly convey. For the bad plot, do the opposite: emphasise what features of the results the plot obscures or makes unclear. In doing so, communicate to me that you understand what makes a good plot to convey the results of quantiative analyses to viewers.

Consider making use of the many extensions to ggplot2, e.g.:

- https://r-graph-gallery.com
- https://krzjoa.github.io/awesome-r-dataviz/#/
- ggdist https://mjskay.github.io/ggdist/
- ggrepel
- ggpubr
- see
- ggsignif
- and others: https://exts.ggplot2.tidyverse.org/gallery/

# Simulate dataset

This code block simulates the data set used for the plots. The researchers who planned and performed this (hypothetical) experiment are investigating the effects of interacting with animals on subsequent human social behavior. Specifically, they hypothesized that interacting with a cute charming cat ahead of a public goods game leads to more prosocial decisions. A within design was used to confirm this hypothesis: Participants were invited two times into the lab (with the appointments being one week apart) where they either had the possibility to play with a cat for 15min (cat condition) or build stone towers for the same amount of time (stone condition). After these 15min, the participants took part in a public goods game with real-life consequences (some trials were chosen randomly and the participants were paid accordingly). Lastly, participants were asked to rate how enjoyable the activity before the game was and indicate their baseline fascination with cats (in the cat condition) or stone tower building (in the stone condition). All in all, the data of 106 participants was acquired.

```{r}

between <- list(condition = c(cat = "Cat Condition", 
                              stone = "Stone Tower Condition"))
               
mu <- data.frame(
  cat    = 20,
  stone  = 23
)
df <- sim_design(between, 
                 n = 106, mu = mu, 
                 sd = 5, r = .5)

```

# Good plot

Do something challenging, but not something so complex that it is useless 
(good example: Raincloud plots)

```{r}



```

# Bad plot

```{r}



```

# Write to disk

Write to disk as a .pdf or .png.

Name them "plot_good_[yourname]" and "plot_bad_[yourname]".

```{r}



```

# Session info

```{r}

sessionInfo()

```


